Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Intuition.tex
probably fixed one comment

\section{Proof that the MSI Protocol Preserves Store Atomicity}
\label{Sec:Safety}
\label{safety}

@@

We first reduce the proof of store atomicity to 4 invariants, presented in
Section~\ref{firstLevel}. These invariants in turn can be reduced to 4 more
invariants presented in Section~\ref{secondLevel}. The reason for choosing the 8
 invariants is because they are not specific to MSI protocol
alone; any other protocol which obeys these invariants obeys Theorem
\fullref{storeAtomicity}. We finally present the key insight which we
discovered for proving the second-level invariants in the form of 3 invariants.
We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI and MOESI also
satisfy these 3 fundamental properties, making their verification easier by reusing
portions of our MSI proof.
We informally show why this is the case using the example of MESI,
MOSI and MOESI in Section~\ref{sec:Refinements}.
\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

@@

We present what we believe to be the key insight for proving any cache-coherence protocol.
In fact, the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address if and only if it sends a response for that address
\item A parent upgrades the directory entry for an address for one of
  its children if and only if it sends a response for that child and
  address
\item The state or directory entry changes for an address only on sending and
receiving a response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}
\vspace{-.1in}

@@

Figure~\ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariant~\ref{stateChange} to \ref{respFIFO}, it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure~\ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant~\fullref{conservative} holds. Since Invariants~\ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section~\ref{secondLevel} requires using more
properties of the MSI transistions of Figure~\ref{trans}, though it is
straightforward. Analysing a parent-child pair will suffice in proving
Invariants~\fullref{localCompatibility1} and
\fullref{localCompatibility2}. Proof of Invariant~\fullref{noTransitWrite} is
similar to the proof of Invariant~\fullref{nonAncestorCompatible}.

>>>>

\section{Proof that the MSI Protocol Preserves Store Atomicity}
\label{safety}

@@

We first reduce the proof of store atomicity to 4 invariants, presented in
Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
 invariants is because they are not specific to the MSI protocol;
any other protocol that obeys these invariants obeys Theorem
\fullref{storeAtomicity}. We finally present the key insight that we
discovered for proving the second-level invariants in the form of 3 invariants.
We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI and MOESI also
satisfy these 3 fundamental properties, making their verification easier by reusing
portions of our MSI proof.
We informally show why this is the case using the examples of MESI,
MOSI and MOESI in Section \ref{sec:Refinements}.
\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

@@

%ndave: This needs some repointing
We present what we believe to be the key insight for proving any cache-coherence protocol.
In fact, the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address.
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address.
\item The state or directory entry changes for an address only on sending and
receiving a response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}

\vspace{-.1in}

@@

Figure \ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure \ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure \ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
\fullref{noTransitWrite} is similar to the proof of Invariant
\fullref{nonAncestorCompatible}.
